---
title: "2017 Boston Marathon Finisher Analysis"
author: "Will Bowden — wbowden@floridapoly.edu"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: true
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 5.5
)
```

## Introduction

The 2017 Boston Marathon attracted **26,410 finishers** from around the world, ranging in age from 18 to 84. This report explores finish-time distributions across gender, age groups, and countries of origin using official race results.

**Data source:** `marathon_results_2017.csv` from [github.com/aalhamadani/Data_vis_rep](https://github.com/aalhamadani/Data_vis_rep).

Key variables:

| Variable | Description |
|---|---|
| `M/F` | Athlete's gender (M / F) |
| `Age` | Athlete's age on race day |
| `Country` | Country of origin (3-letter code) |
| `Official Time` | Chip-timed finish time (hh:mm:ss) |
| `Half` | First-half split time |

---

## Data Loading & Wrangling

```{r load-data}
library(tidyverse)
library(plotly)

marathon <- read_csv("../data/marathon_results_2017.csv", col_types = cols())

marathon <- marathon |>
  mutate(
    time_hrs = as.numeric(`Official Time`) / 3600,
    gender   = if_else(`M/F` == "M", "Male", "Female"),
    age_group = cut(
      Age,
      breaks = c(17, 24, 34, 44, 54, 64, Inf),
      labels = c("18–24", "25–34", "35–44", "45–54", "55–64", "65+")
    )
  ) |>
  filter(!is.na(time_hrs))

glimpse(marathon)
```

### Summary statistics by gender

```{r summary-by-gender}
marathon |>
  group_by(gender) |>
  summarise(
    n        = n(),
    median_h = round(median(time_hrs), 2),
    mean_h   = round(mean(time_hrs), 2),
    sd_h     = round(sd(time_hrs), 2),
    min_h    = round(min(time_hrs), 2),
    max_h    = round(max(time_hrs), 2)
  ) |>
  knitr::kable(caption = "Finish-time summary by gender")
```

---

## Visualization 1 — Interactive Finish-Time Distributions by Gender

Hover over any bar to see exact density values. The density curves show that both groups are right-skewed — most finishers cluster near 3–4 hours with a longer tail toward slower times. Female finishers peak slightly later (~4.1 h) than male finishers (~3.7 h).

> **Interactivity:** Hovering reveals exact density values at any point on the curve. Clicking a legend entry toggles that group on/off, allowing isolated inspection of either gender's distribution.

```{r plot-density-interactive, fig.alt="Interactive density plot of finish times for male and female runners. Both distributions are right-skewed. Male finishers peak near 3.7 hours; female finishers peak near 4.1 hours. Dashed lines mark group medians."}
medians <- marathon |>
  group_by(gender) |>
  summarise(med = median(time_hrs))

# Build the density data manually for plotly
dens_f <- density(marathon$time_hrs[marathon$gender == "Female"], from = 2, to = 8)
dens_m <- density(marathon$time_hrs[marathon$gender == "Male"],   from = 2, to = 8)

plot_ly() |>
  add_trace(
    x = dens_f$x, y = dens_f$y,
    type = "scatter", mode = "lines",
    fill = "tozeroy",
    fillcolor = "rgba(230,159,0,0.3)",
    line = list(color = "#E69F00", width = 2),
    name = "Female",
    hovertemplate = "Time: %{x:.2f} h<br>Density: %{y:.4f}<extra>Female</extra>"
  ) |>
  add_trace(
    x = dens_m$x, y = dens_m$y,
    type = "scatter", mode = "lines",
    fill = "tozeroy",
    fillcolor = "rgba(0,114,178,0.3)",
    line = list(color = "#0072B2", width = 2),
    name = "Male",
    hovertemplate = "Time: %{x:.2f} h<br>Density: %{y:.4f}<extra>Male</extra>"
  ) |>
  add_segments(
    x = medians$med[medians$gender == "Female"],
    xend = medians$med[medians$gender == "Female"],
    y = 0, yend = 0.55,
    line = list(color = "#E69F00", dash = "dash", width = 1.5),
    name = "Female median", showlegend = FALSE,
    hovertemplate = paste0("Female median: ", round(medians$med[medians$gender == "Female"], 2), " h<extra></extra>")
  ) |>
  add_segments(
    x = medians$med[medians$gender == "Male"],
    xend = medians$med[medians$gender == "Male"],
    y = 0, yend = 0.55,
    line = list(color = "#0072B2", dash = "dash", width = 1.5),
    name = "Male median", showlegend = FALSE,
    hovertemplate = paste0("Male median: ", round(medians$med[medians$gender == "Male"], 2), " h<extra></extra>")
  ) |>
  layout(
    title = list(
      text = "Finish-Time Distribution — 2017 Boston Marathon<br><sup>Dashed lines mark group medians · hover to explore</sup>",
      font = list(size = 16)
    ),
    xaxis = list(title = "Official Finish Time (hours)", tickvals = 2:8,
                 ticktext = paste0(2:8, " h")),
    yaxis = list(title = "Density"),
    legend = list(orientation = "h", y = -0.15),
    hovermode = "x unified"
  )
```

**Takeaway:** The median male finish time is approximately 3.7 h and the median female finish time is approximately 4.1 h — a roughly 30-minute gap. Both distributions are right-skewed, reflecting that the slowest finishers pull the tails out further than the fastest pull in the other direction.

---

## Visualization 2 — Finish Times by Age Group & Gender

Box plots show how finishing speed changes with age. Peak performance appears in the 25–34 bracket; times generally worsen through older brackets. The 18–24 group is not the fastest, likely reflecting less race experience among younger entrants.

```{r plot-boxplot, fig.alt="Box plots of marathon finish times grouped by age bracket and colored by gender. The 25-34 age group has the lowest median finish times for both genders. Times increase with age, and spread widens for older runners."}
ggplot(marathon, aes(x = age_group, y = time_hrs, fill = gender)) +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.3, linewidth = 0.5) +
  scale_y_continuous(
    breaks = 2:8,
    labels = function(x) paste0(x, " h")
  ) +
  # Okabe-Ito colorblind-safe palette
  scale_fill_manual(values = c("Female" = "#E69F00", "Male" = "#0072B2")) +
  labs(
    title    = "Finish Time by Age Group and Gender",
    subtitle = "2017 Boston Marathon (n = 26,410)",
    x        = "Age Group",
    y        = "Official Finish Time",
    fill     = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

---

## Visualization 3 — Top Countries by Number of Finishers

The USA dominates participation. The chart below focuses on the **top 15 countries excluding the USA** to highlight international diversity. Canada and Great Britain lead the international field.

```{r plot-countries, fig.alt="Horizontal stacked bar chart of the top 15 non-US countries at the 2017 Boston Marathon. Canada leads with roughly 850 finishers, followed by Great Britain and Mexico. Bars are split by gender using orange for female and blue for male."}
top_countries <- marathon |>
  filter(Country != "USA") |>
  count(Country, gender) |>
  group_by(Country) |>
  mutate(total = sum(n)) |>
  ungroup() |>
  slice_max(total, n = 30) |>
  mutate(Country = fct_reorder(Country, total))

ggplot(top_countries, aes(x = n, y = Country, fill = gender)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("Female" = "#E69F00", "Male" = "#0072B2")) +
  labs(
    title    = "Top 15 Non-US Countries at the 2017 Boston Marathon",
    subtitle = "Stacked by gender",
    x        = "Number of Finishers",
    y        = NULL,
    fill     = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

---

## Visualization 4 — Second-Half Fade by Gender

A classic marathon challenge is "the fade": runners slow in the second half. Positive fade values mean the runner took longer in the second half than the first.

```{r plot-fade, fig.alt="Density plot of second-half fade (second half minus first half, in minutes) for male and female runners. Most runners are positive (slowing down). Female runners show a slightly tighter distribution than males, suggesting more consistent pacing."}
marathon_fade <- marathon |>
  mutate(
    half_sec    = as.numeric(Half),
    second_half = time_hrs * 3600 - half_sec,
    fade_min    = (second_half - half_sec) / 60
  ) |>
  filter(!is.na(fade_min), abs(fade_min) < 120)

ggplot(marathon_fade, aes(x = fade_min, fill = gender, color = gender)) +
  geom_density(alpha = 0.35, linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey40") +
  scale_fill_manual(values  = c("Female" = "#E69F00", "Male" = "#0072B2")) +
  scale_color_manual(values = c("Female" = "#E69F00", "Male" = "#0072B2")) +
  labs(
    title    = "Second-Half Fade at the 2017 Boston Marathon",
    subtitle = "Positive values = runner slowed in second half",
    x        = "Second Half − First Half (minutes)",
    y        = "Density",
    fill = NULL, color = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

**Takeaway:** Almost all runners slow in the second half (peak of the distribution is positive). Female runners show a slightly tighter, more concentrated fade distribution than males, consistent with more conservative early-race pacing.

---

## Discussion

### What were the original charts you planned to create for this assignment?

I originally planned a finish-time histogram by gender, an age-vs-time scatter plot, and a country participation bar chart. After exploring the data, I refined the approach:

- Density plots convey distributional shape more clearly than histograms for 26,000+ observations.
- Box plots grouped by age bracket communicate the age-performance relationship more compactly than a raw scatter.
- The second-half fade analysis surfaced an interesting behavioral pattern that a country bar chart would have missed.

### What story could you tell with your plots?

1. **Gender gap** — Male and female finishers have distinct distributions, separated by ~30 minutes at the median.
2. **Age and performance** — Peak marathon performance occurs in the 25–44 bracket; both younger and older runners tend to run slower, though older brackets show wider spread.
3. **Global reach** — While ~79% of finishers were American, the race draws runners from dozens of countries, led internationally by Canada and Great Britain.
4. **Pacing behavior** — Almost all runners slow in the second half; female runners' tighter fade distribution suggests more conservative early-race strategy.

### How did you apply the principles of data visualization and design?

- **Accessibility:** All figures use the Okabe-Ito colorblind-safe palette (orange `#E69F00` for Female, blue `#0072B2` for Male). Alt text is provided for every figure via `fig.alt`. Gender is also encoded in the legend label so color alone is not the only channel.
- **Interactivity (Plot 1):** The finish-time density chart is rendered with `plotly`, allowing hovering for exact values and toggling gender groups on/off — impossible in a static image.
- **Reduce chartjunk:** `theme_minimal()` throughout removes unnecessary grid lines and backgrounds.
- **Appropriate geometry:** Density and box plots manage overplotting with 26,000+ observations.
- **Informative labels:** Every plot has a descriptive title, subtitle, and axis labels.
